---
title: "Practical Machine Learning Course Project"
output: html_document
---

## Overview

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, the goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict the manner in which they did the exercise as they were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

The training data for this project are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har 


## Loading and Cleaning the Data

First, we load the data and perform the necessary transformations to set the data up for analysis.

```{r warning = FALSE, message=FALSE}
## Set Environment
library(knitr)
library(gbm)
library(caret)
library(rpart)
library(randomForest)
set.seed(12345)

## Read Data
trainSet <- read.csv("pml-training.csv")
testSet <- read.csv("pml-testing.csv")
str(trainSet)
str(testSet)

## Partition Training Set
toPart <- createDataPartition(trainSet$classe, p = 0.7, list = FALSE)
partTrain <- trainSet[toPart, ]
partTest <- trainSet[-toPart, ]

## Cleaning Transformation 1: Removing Near Zero Variance
clean1 <- nearZeroVar(partTrain)
partTrain <- partTrain[, -clean1]
partTest <- partTest[, -clean1]

## Cleaning Transformation 2: Removing NAs
clean2 <- sapply(partTrain, function(x) mean(is.na(x))) > 0.95
partTrain <- partTrain[, clean2 == FALSE]
partTest <- partTest[, clean2 == FALSE]

## Cleaning Transformation 3: Removing Unnecesary Variables (IDs)
partTrain <- partTrain[, -(1:5)]
partTest <- partTest[, -(1:5)]

## Results
dim(partTrain)
dim(partTest)
```


## Building the Models

Second, we design 3 models to cross-validate our in sample test data and compare their results to decide which one works best.

```{r warning = FALSE}
## Decision Tree
dtMod <- rpart(classe ~ ., data = partTrain, method ="class")
dtModPredict <- predict(dtMod, partTest, type ="class")
dtModCMat <- confusionMatrix(dtModPredict, partTest$classe)
dtModAccuracy <- dtModCMat$overall['Accuracy']

## Generalized Boosting
gbTrControl <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
gbMod <- train(classe ~ ., data = partTrain, method = "gbm", trControl = gbTrControl, verbose = FALSE)
gbModPredict <- predict(gbMod, partTest)
gbModCMat <- confusionMatrix(gbModPredict, partTest$classe)
gbModAccuracy <- gbModCMat$overall['Accuracy']

## Random Forest
rfTrControl <- trainControl(method = "cv", number=3, verboseIter = FALSE)
rfMod <- train(classe ~ ., data = partTrain, method ="rf", trControl = rfTrControl)
rfModPredict <- predict(rfMod, partTest)
rfModCMat <- confusionMatrix(rfModPredict, partTest$classe)
rfModAccuracy <- rfModCMat$overall['Accuracy']

## Results
resultsMat <- cbind(dtModAccuracy, gbModAccuracy, rfModAccuracy)
resultsMat
```

The Random Forest Model proved to be the most accurate at 99.66%. With this model, the expected out of sample error would be 0.34%. Due to these great results, this model will be selected to use on the Test Set.


## Applying Model to Test Set

Thirdly, we apply the chosen model on the Test Set and predict the class of the 20 cases.

```{r warning = FALSE}
predictTEST <- predict(rfMod, testSet)
predictTEST
```



